---
title: "Review of basic statistical concepts"
subtitle: "Ecological Data Management and Analysis"
author: "Achaz von Hardenberg"
date: "10 Nov 2014"
output:
  ioslides_presentation:
    logo: eco_wordle.jpg
    widescreen: true
    incremental: true
---  

## Introducing the Dahu
  
<iframe width="420" height="315" src="//www.youtube.com/embed/BBMVraJZoAU" frameborder="0" allowfullscreen></iframe>  
   
## Ok, we have got the data... and now? {.build}

Measures of **Central value:**

*Mean*  

*Median*

*Mode*

## The Mean {.build}

$\large{\frac{\sum{x}}{n}}$    

```{r}
dahu.dat<-read.csv("dahu.csv")
dahuM.dat<-subset(dahu.dat,Sex==1)
```

```{r}
meanMR<-sum(dahuM.dat$R)/length(dahuM.dat$R)
```

```{r}
meanMR
``` 
But of course there must be a function for this, right?  
```{r}
mean(dahuM.dat$R)
```

## The Median {.build}  

The Median is the value dividing the sample exactly in half (in a sorted list of numbers).

```{r}
dahuM.dat<-dahuM.dat[order(dahuM.dat$R),]
```
 
```{r}
length(dahuM.dat$R)/2
```

```{r}
dahuM.dat$R[34]
```

```{r}
median(dahuM.dat$R)
```

## The Normal distribution {.build} 

The sample mean ($\overline{x} = 34.75$) and the median ($\tilde{x} = 35.13$) look very similar... why is that?  

```{r}
hist(dahuM.dat$R)
```

## The Normal distribution

The sample mean ($\overline{x} = 34.75$) and the median ($\tilde{x} = 35.13$) look very similar... why is that?  

```{r}
hist(dahuM.dat$R,prob=T)
```


## The Normal distribution

```{r}
hist(dahuM.dat$R,prob=T)
lines(density(rnorm(1000000,mean(dahuM.dat$R),sd(dahuM.dat$R))),col="red")
```

## The Normal distribution

```{r}
hist(dahuM.dat$R,prob=T)
lines(density(rnorm(1000000,mean(dahuM.dat$R),sd(dahuM.dat$R))),col="red")
segments(mean(dahuM.dat$R),0,mean(dahuM.dat$R),0.2,col="blue")
```

## The Normal distribution

```{r}
hist(dahuM.dat$R,prob=T)
lines(density(rnorm(1000000,mean(dahuM.dat$R),sd(dahuM.dat$R))),col="red")
segments(mean(dahuM.dat$R),0,mean(dahuM.dat$R),0.2,col="blue")
segments(median(dahuM.dat$R),0,median(dahuM.dat$R),0.2,col="green")
```

## The Mode {.build}

The mode is the number which appears most often in a sample (works with integers)
```{r}
x <- c(1,2,2,2,3,4,5,6,7,8,9)
```

```{r}
temp <- table(as.vector(x))
```

```{r}
names(temp)[temp == max(temp)]
```

## The Mode {.build}  

For continous data in `R`: 
```{r}
library(modeest)
mlv(dahuM.dat$R,method="mfv")
```


