---
title: "Probability in a Nutshell"
subtitle: "Ecological Data Management and Analysis"
author: "Achaz von Hardenberg"
date: "3 Nov 2014"
output:
  ioslides_presentation:
    logo: eco_wordle.jpg
    widescreen: true
    incremental: true
---

## What is probabilty? {.build}

Probability theory is the mathematical abstraction formalizing the mathematics of uncertain events

In Southern Chile sheep can be predated by Culpeo foxes, pumas or stray dogs:
  
Predation by puma is an **event**  
  
to be predated or not is an **outcome** of the above event
  
**sample space**: the set of all possible outcomes of a specific event: ex. sheep predated by fox, puma, dog or the probability of a coin being flipped head or tail
 
## What is the probabilty that flipping a coin it comes out "head"? {.build}
 
One common way to interpret *probability* is as the  
  
*long run frequency of an outcome in a series of repeated trials*

Which means: the probability that the coin comes out head is equal to the proportion of times it comes out "head", repeating the coin flipping experiment infinite times (frequency interpretation of probability).

## Running proportion of heads flipping a coin

trial:      
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(700,1,0.5)
  trial<-(0)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial: 1    
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(700,1,0.5)
  trial<-(1)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial: 2    
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(700,1,0.5)
  trial<-(1:2)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:3        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(3,1,0.5)
  trial<-(1:3)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:4        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(4,1,0.5)
  trial<-(1:4)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:5        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(5,1,0.5)
  trial<-(1:5)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:6        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(6,1,0.5)
  trial<-(1:6)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:7        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(7,1,0.5)
  trial<-(1:7)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:8        
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(8,1,0.5)
  trial<-(1:8)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:9       
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(9,1,0.5)
  trial<-(1:9)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:10  
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(700,1,0.5)
  trial<-(1:10)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:50         
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(50,1,0.5)
  trial<-(1:50)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:100   
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(100,1,0.5)
  trial<-(1:100)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Running proportion of heads flipping a coin
trial:500    
```{r echo=FALSE }
  set.seed(4564563)
  heads<-rbinom(500,1,0.5)
  trial<-(1:500)
  heads.df<-data.frame(trial,heads)
  heads.df <- within(heads.df, prop <- cumsum(heads)/trial)
  plot(heads.df$prop~log(heads.df$trial),ylim=c(0,1),xlim=c(0,7),xlab="Trial number (log scale)",ylab="Proportion of heads")
  cord<-xy.coords(-100:700,0.5:0.5,recycle=T)
  lines(cord,type="l",col="red")
```

## Basic probability laws {.build}

Mathematical laws ruling the the combination of  *events*

**Mutually exclusive events**

Predation by Culpeo fox or by puma are mutually exclusive events (a single sheep can not be predated by both)

**Sheep Predation Probabilities by different predators**

Symbol|Species|Probability|
-|-----|-----|
F|Fox  |0.026|
P|Puma |0.010|
D|Dog  |0.063|

## Basic probability laws {.build}
 
What is the probability to be predated by a fox (F) *or* by a puma (P)?  

$P(F \cup P) = P(F)+P(P)$     

The $\cup$ symbol means 'Union' 

$P(Fox \ or \ Puma) = 0.026+0.010$

$P(Fox \ or \ Puma) = 0.036$

What is the probability to be predated by any predator?

$P(Fox \ or \ Puma \ or \ Dog) = P(Fox)+P(Puma)+P(Dog)$

$P(Fox \ or \ Puma \ or \ Dog) = 0.026+0.010+0.063 = `r 0.026+0.010+0.063`$

## Non-mutually exclusive events {.build}

Sheep can be white or black  
The color (C) of the sheep is the *event* and black or white are the two possible  *outcomes*

What is the probabilty that a sheep is black?  

Let's say it's $P(C_{black}) = 0.2$

> **The probabilities of all the possible outcomes sum up to 1**

What is the probabilty that a sheep is white?

$P(C_{white}) = 1 - P(C_{black}) = 0.8$

## Non-mutually exclusive events {.build}

What is the probabilty that a sheep is of a particular color **or** is predated by a stray dog? 

$P(C \cup D) = P(C)+P(D)-P(C \cap D)$   

$P(C \cap D)$ is the *joint* probabilty (intersection) of being both of a particular color and being predated by a stray dog. If the color of the sheep and predation by dogs are *independent* events (i.e. knowing about one event does not provide information about the other) then: 

$P(C \cap D) = P(C)P(D)$  

and thus

Probability of being Black or Predated by Dog = 0.2+0.063-0.2x0.063 =`r 0.2+0.063-0.2*0.063`

Probability of being White or Predated by Dog = 0.8+0.063-0.8x0.063 =`r 0.8+0.063-0.8*0.063`

## Conditional probability {.build}
 
Let's say we visit 100 alpine lakes and record if there is the presence of alien brook trout (*Salvelinus fontinalis*)  and of common frogs (*Rana temporaria*):  
  
Event F = Frog detection: probability to detect a frog during a visit   
Event T = Trout presence: probability that a lake is inhabited by alien brook trout 

/|Trouts|no Trouts|Total
-|-----|-----|----
Frogs|5  |35|40
No Frogs|45 |15|60
Total|50|50|100

$P(T) = 50/100 = 0.5$  
$P(F) = 40/100 = 0.4$ 
$P(F \cap T) = 5/100 = 0.05$  

## Conditional Probability {.build}

What is the probability to find common frogs (*Rana temporaria*) in Alpine lakes in which there is the presence of brook trout?

$P(F|T) = \frac{P(F \cap T)}{P(T)}$

$P(F|T) =  0.05/0.5 = 0.1$

So... what is the conditional probability to find frogs in lakes without brook trout?  

$P(F|T_{no}) =  0.35/0.5 = 0.7$


## Bayes' Theorem {.build}

if   
$P(A|B) = \frac{P(A \cap B)}{P(B)}$

we can also write  
$P(A \cap B) = P(A|B)P(B)$

and   
$P(B \cap A) = P(B|A)P(A)$

and because  
$P(A \cap B) = P(B \cap B)$  
$P(B|A)P(A) = P(A|B)P(B)$

re-arranging we get the **Bayes' Theorem**:  
  
$$\large{P(B|A) = \frac{P(A|B)P(B)}{P(A)}}$$

## Bayes' Theorem {.build}

What is the probability that there are brook trout in a lake if I found frogs? 

$P(T|F) = \frac{P(F|T)P(T)}{P(F)}$

$P(T|F) = \frac{0.1 \times 0.5}{0.4} = 0.125$












